Composable Generalized Consensus

### 2016-01-14 paper/cgc.tex
Work on the paper, with the goal of presenting it to colleagues.

We consider a computing service specified as a deterministic state machine $M$ whose transitions correspond to the commands that clients of the service may issue.
If $M$ is implemented on a single server then any fault happening on this server may become visible to the clients of the service.
To mask faults, one can replicate the implementation of $M$ on several replica servers (called replicas for short) and direct clients' commands to replicas which are known to be non-faulty.
However, the intent of replication is to mask faults while preserving the behavior of the service from the point of view of its clients.

@@

Generalized Consensus~\cite{Lamport05GeneralizeConsensus} is a formal specification of the requirements that the distributed algorithm $A$ must meet in order for SMR to provide the illusion of a centralized fault-tolerant service.
Generalized Consensus uses the notion of a set of command-structures, or c-structs. C-structs are obtained by appending commands to the c-struct $\bot$. Given a c-struct $s$ and a command $c$, $s \bullet c$ is the c-struct obtained by appending $c$ to $s$. Given a sequence of commands $cs$, $s\star cs$ is the c-struct obtained by repeatedly appending the commands in $cs$ in the order where they appear in $cs$. By appropriately defining the operator $\bullet$ according to the
definition of $M$, C-structs can describe sequences of commands up the reordering of commuting
commands: it is possible to have $s \star \left[c_1,c_2\right] = s \star \left[c_2,c_1\right]$.
Moreover, c-structs must satisfy a number of properties: \setword{property 1}{prop:1}, c-structs must form a partial order under the operator $\preceq$ defined as: $s_1 \preceq s_2$ if and only if there exists a sequence of commands $cs$ such that $s_2 = s_1 \star cs$; \setword{property 2}{prop:2}, the greatest lower bound of a set of c-structs $S$ must always exists and be constructible from the same commands as the c-structs in $S$; and \setword{property 3}{prop:3}, a set of c-structs $S$
which has an upper bound, called a \emph{compatible} set, must have a lowest upper bound which is constructible from the same commands as the c-structs in $S$.

Generalized Consensus is formulated in terms of two types of processes: the set $P$ of proposers and the set $L$ of learners, but it is possible for each replica to play the two roles at the same time. Proposers propose commands and learners, which represent replicas, learn about c-structs.
We describe the execution of learners and proposers using the variable $prop$, which is a set containing all the commands that have been proposed, and the variable $learned$, which maps a learner to the latest c-structs it has learned.

Generalized Consensus requires the following properties: \emph{non-triviality}, stating that for any learner $l$, there is at all times a sequence of proposed commands $cs$ such that $learned\left[l\right] = \bot \bullet cs$; \emph{stability}, stating that for any learner $l$, $learned\left[ l \right]$ increase with time; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always \emph{compatible}; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$.
Note that the FLP theorem~\cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess} implies that the liveness property cannot be achieved in a networked system where processes may fail. Therefore, the liveness property guaranteed by Generalized Consensus algorithms is necessary weaker that the liveness property state above, maybe adding assumptions about faults and synchrony.
We refer the reader to the work of Lamport~\cite{Lamport05GeneralizeConsensus} for a thorough discussion of Generalized Consensus.

@@

Composable Generalized Consensus augments the Generalized Consensus specification with an abort interface that allows an SMR module to abort its execution, and with an init interface which allows an SMR module to start its execution where a preceding aborted module left off.
To specify the composition interface, we introduced a third type of processes called the \emph{switchers}. Switchers can receive \emph{init c-structs} from a preceding aborting modules and can pass \emph{abort c-structs} to pass the baton to a next SMR module when the current SMR module aborts.

We describe the behavior of switchers using the variable $from$, which contains all the init c-structs received by any switcher, and the variable $to$, which contains all the abort c-structs produced by the switchers.
The variable $prop$ is used as before, but the variable $learned$ now maps learners to sets of c-structs and, for each learner $l$, $learned\left[ l \right]$ consists of all the c-structs learned so far by $l$.
We define the set of valid c-structs as the c-structs of the form $t\star cs$ where $cs$ is a sequence of proposed commands, and $t$ is the greatest lower bound of a non-empty subset of the init c-structs received.

Composable Generalized Consensus requires the following properties: \emph{initialization}, stating that no value be learned unless at least one init c-struct have been received; \emph{non-triviality}, stating that for any learner $l$, $learned\left[ l \right]$ is valid; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always compatible; \emph{safe-abort}, stating that every abort c-struct is valid, that for any abort c-struct $s_a$ and learned value $s_l$ we have
$s_l \preceq s_a$; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$ or a switcher eventually produces an abort c-struct.
A TLA$^+$ specification of the safety properties of CGC appears in \cref{fig:cgc}.

We say that an algorithm implementing CGC under some refinement mapping is a \emph{CGC instance}, and that two CGC instances $i_1$ and $i_2$ are compatible if they update disjoint variables except for the $to$ variable of $i_1$ and the $from$ variable of $i_2$, which must be identical.
Note that in the execution of two compatible instances, any abort c-struct produced by the first instance is an init c-struct for the second instance.
We say that a CGC instance is an initial instance when its set of init c-structs is always a subset of $\left\{ \bot \right\}$.

CGC enjoys two main properties: first, the composition of two CGC instances is itself a CGC instance when the first instance is an initial instance; second, an initial CGC instance satisfies the safety properties of Generalized Consensus under the restrictions explained below.
We call the first property the \emph{composition theorem}.
A formal statement of the composition theorem appears in \cref{fig:compthm}, and an informal proof appears in \cref{sec:compproof}. 
Note that TLA$^+$ is a second order logic and can therefore not express the composition theorem as stated above because we cannot quantify of modules.
Instead, we state the composition theorem for two CGC instances obtained by only renaming the variables of the CGC specification.
Any CGC instance would trivially refine both of the two instances used in formalization of the composition theorem because a CGC instance refines the CGC specification.
Therefore, we obtain the composition theorem by transitivity and monotonicity of refinement~\cite{AbadiLamport91ExistenceRefinementMappings}.

@@

\subsection{Proof of the Composition Theorem}%
\label{sec:compproof}
To prove the composition theorem, as expressed in \cref{fig:compthm}, we must show that $Init$ implies $CGC!Init$ and that every transition of $Spec \equiv Init \wedge \box\left[Next\right]_{vars}$ is a transition of $CGC!Spec$.
Below we sketch a proof of the second condition.

@@

Similarly, the $CGC2!Learn(l)$ transition inserts a valid c-struct $s$ in the set $CGC!learned\left[ l \right]$, and we must show the same 3 properties.
Property (1) is established by the first conjunct of the $CGC2!Learn\left( l \right)$ transition, using the fact that $CGC!to = CGC2!to$.
To show (2), consider a learned c-struct of $CGC$.
There are two cases: (2a), $s_l$ is a learned c-struct of $CGC1$, and (2b), $s_l$ is a learned c-struct of $CGC2$. In case (2b), the second conjunct of the $CGC2!Learn(l)$ transition establishes that $s$ and $s_l$ are compatible. In case (2a), by reasoning similarly as in TODO use an invariant...
To show (3)\ldots
TODO\@: complete.
We have formalized informal proof above in Isabelle/HOL (and not TLAPS~\cite{CousineauETAL12TlaProofs}, because the authors are proficient in Isabelle/HOL but not in TLAPS); the proof is available at \url{http://losa.fr/cgc}.

@@

>>>>

We consider a computing service specified as a deterministic state machine $M$ whose transitions correspond to the commands that clients of the service may issue.
If $M$ is implemented on a single server then any fault happening on this server may become visible to the clients of the service.
To mask faults, one can replicate the implementation of $M$ on several replica servers (called replicas for short) and direct clients' commands to replicas which are known to be non-faulty.
However, the intent of replication is to mask faults while preserving the behavior of the service from the point of view of its clients.
Therefore, replicas must be somehow synchronized to act as a centralized implementation.

@@

Composable Generalized Consensus is based on Generalized Consensus, and any CGC algorithm can be used as a Generalized Consensus algorithm. Therefore, we start by a review of generalized and present CGC in the next subsection.

Generalized Consensus~\cite{Lamport05GeneralizeConsensus} is a formal specification of the requirements that the distributed algorithm $A$ described above must meet in order for SMR to provide the illusion of a centralized fault-tolerant service.
Generalized Consensus uses the notion of command-structures, or c-structs. C-structs are obtained by appending commands to the c-struct $\bot$. Given a c-struct $s$ and a command $c$, $s \bullet c$ is the c-struct obtained by appending $c$ to $s$. Given a sequence of commands $cs$, $s\star cs$ is the c-struct obtained by repeatedly appending the commands in $cs$ in the order where they appear in $cs$. By appropriately defining the operator $\bullet$ according to the
definition of the state-machine $M$, c-structs can stand for sequences of commands up the reordering of commuting
commands; for example, we can have $s \star \left[c_1,c_2\right] = s \star \left[c_2,c_1\right]$ for some c-struct $s$ and two commands $c_1$ and $c_2$.

C-structs must satisfy a number of properties: \setword{property 1}{prop:1}, c-structs must form a partial order under the operator $\preceq$ defined as: $s_1 \preceq s_2$ if and only if there exists a sequence of commands $cs$ such that $s_2 = s_1 \star cs$; \setword{property 2}{prop:2}, the greatest lower bound of a set of c-structs $S$ must always exist and be constructible from the same commands as the c-structs in $S$; and \setword{property 3}{prop:3}, a set of c-structs $S$
which has an upper bound, called a \emph{compatible} set, must have a least upper bound which is constructible from the same commands as the c-structs in $S$.

Generalized Consensus is formulated in terms of two types of processes: the set $P$ of proposers and the set $L$ of learners, but it is possible for each replica to play the two roles at the same time. Proposers propose commands and learners, which represent replicas, learn about c-structs.

We specify the execution of learners and proposers in terms of a transition system consisting of the variable $prop$, which is a set containing all the commands that have been proposed, and the variable $learned$, which maps a learner to the latest c-structs it has learned.

Generalized Consensus requires the following properties: \emph{non-triviality}, stating that for any learner $l$, there is at all times a sequence of proposed commands $cs$ such that $learned\left[l\right] = \bot \bullet cs$; \setword{\emph{stability}}{prop:stability}, stating that for any learner $l$, $learned\left[ l \right]$ increase with time; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always \emph{compatible}; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$.
Note that the FLP theorem~\cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess} implies that the liveness property cannot be achieved in a networked system where processes may fail. Therefore, the liveness property guaranteed by Generalized Consensus algorithms is necessary weaker that the liveness property state above, maybe adding assumptions about faults and synchrony.
We refer the reader to the work of Lamport~\cite{Lamport05GeneralizeConsensus} for a thorough discussion of Generalized Consensus.

@@

Composable Generalized Consensus augments the Generalized Consensus specification with an abort interface that allows an SMR module to abort its execution, and with an init interface which allows an SMR module to start its execution where a preceding aborted module left off.
To specify the composition interface, we introduce a third type of processes called the \emph{switchers}. Switchers can receive \emph{init c-structs} from a preceding aborting module and can pass \emph{abort c-structs} to a next SMR module, which will treat them as init c-structs, when the current SMR module aborts.

We describe the behavior of switchers using the variable $from$, which contains all the init c-structs received by the switchers, and the variable $to$, which contains all the abort c-structs produced by the switchers.
The variable $prop$ is used as before, but the variable $learned$ now maps learners to sets of c-structs and, for each learner $l$, $learned\left[ l \right]$ consists of all the c-structs learned so far by $l$.
We define the set of valid c-structs as the c-structs of the form $t\star cs$ where $cs$ is a sequence of proposed commands, and $t$ is the greatest lower bound of a non-empty subset of the init c-structs received.

Composable Generalized Consensus requires the following properties: \emph{initialization}, stating that no value be learned unless at least one init c-struct has been received; \emph{non-triviality}, stating that for any learner $l$, $learned\left[ l \right]$ is valid; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always compatible; \emph{safe-abort}, stating that every abort c-struct is valid, and that for any abort c-struct $s_a$ and learned value $s_l$ we have
$s_l \preceq s_a$; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$ or a switcher eventually produces an abort c-struct.
A TLA$^+$ specification of the safety properties of CGC appears in \cref{fig:cgc}.

We say that an algorithm implementing CGC under some refinement mapping is a \emph{CGC instance}, and that two CGC instances $i_1$ and $i_2$ are compatible if they update disjoint variables except for the $to$ variable of $i_1$ and the $from$ variable of $i_2$ which must be identical, therefore identifying the abort c-struct of $i_1$ with the init c-structs of $i_2$.
%Note that in the execution of two compatible instances, any abort c-struct produced by the first instance is an init c-struct for the second instance.
We say that a CGC instance is an initial instance when its set of init c-structs is always a subset of $\left\{ \bot \right\}$.

CGC enjoys two main properties: first, the composition of two CGC instances is itself a CGC instance if the first instance is an initial instance; second, an initial CGC instance satisfies the safety properties of Generalized Consensus under the restrictions explained below.
We call the first property the \emph{composition theorem}.
A formal statement of the composition theorem appears in \cref{fig:compthm}, and an informal proof appears in \cref{sec:compproof}. 
Note that TLA$^+$ is a second order logic and can therefore not express the composition theorem as stated above because we cannot quantify of modules.
Instead, we state the composition theorem for two CGC instances obtained by only renaming the variables of the CGC specification.
Any CGC instance would trivially refine both of the two instances used in the formalization of the composition theorem.
Therefore, we obtain the composition theorem by transitivity and monotonicity of refinement~\cite{AbadiLamport91ExistenceRefinementMappings}.
To show that an initial instance implements Generalized Consensus, we must first wrap it in a small layer that prevents the learned c-struct of a learner from changing non-monotonically. This is because the Generalized Consensus specification requires the learned c-struct of a learner to grow monotonically (the \ref{prop:stability} property), whereas CGC allows the learned c-struct of a learner change non-monotonically. However, this is not a fundamental difference: when a learner
learns a new c-struct in CGC, the additional layer simply updates the learned c-struct of a learner to the least upper-bound of its previous value and the value learned in CCG, thereby guaranteeing the stability property of Generalized Consensus.

@@

\subsection{Proof of the Composition Theorem}%
\label{sec:compproof}
\begin{comment}
To prove the composition theorem, as expressed in \cref{fig:compthm}, we must show that $Init$ implies $CGC!Init$ and that every transition of $Spec \equiv Init \wedge \box\left[Next\right]_{vars}$ is a transition of $CGC!Spec$.
Below we sketch a proof of the second condition.

@@

Similarly, the $CGC2!Learn(l)$ transition inserts a valid c-struct $s$ in the set $CGC!learned\left[ l \right]$, and we must show the same 3 properties.
Property (1) is established by the first conjunct of the $CGC2!Learn\left( l \right)$ transition, using the fact that $CGC!to = CGC2!to$.
To show (2), consider a learned c-struct of $CGC$.
There are two cases: (2a), $s_l$ is a learned c-struct of $CGC1$, and (2b), $s_l$ is a learned c-struct of $CGC2$. In case (2b), the second conjunct of the $CGC2!Learn(l)$ transition establishes that $s$ and $s_l$ are compatible. In case (2a), by reasoning similarly as in TODO use an invariant.
To show (3)\ldots
TODO\@: complete.
We have formalized informal proof above in Isabelle/HOL (and not TLAPS~\cite{CousineauETAL12TlaProofs}, because the authors are proficient in Isabelle/HOL but not in TLAPS); the proof is available at \url{http://losa.fr/cgc}.

@@

\end{comment}

### 2016-01-14 paper/intro.tex
Work on the paper, with the goal of presenting it to colleagues.

Surprisingly, at a first glance, many SMR algorithms seem to share a
common structure which would indicate that previous formal analysis could
be reused. That is because, in many cases, researchers obtained a new SMR
algorithm by optimizing an existing SMR algorithm for a particular use
case. For example, Fast Paxos~\cite{Lamport06FastPaxos} adds fast rounds
to MultiPaxos~\cite{lamport2001paxos} to reduce the latency under low
contention, Disk Paxos~\cite{GafniLamport03DiskPaxos} optimizes Paxos
for systems where disk nodes and processor nodes are distinct, Chain
Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability} 
organizes the replicas of Paxos in a ring to distribute
commands at a higher throughput but at the expense of latency,
FGGC~\cite{SutraShapiro11FastGenuineGeneralizedConsensus} optimizes
Generalized Paxos\cite{Lamport05GeneralizeConsensus} to reduce latency
when the system contains a reliable and fast leader node, Ring
Paxos~\cite{MarandiETAL10RingPaxosHighthroughputAtomicBroadcastProtocol}
optimizes the latency and throughput of Paxos when network multicast is
available, Multicoordinated Paxos reduces the performance impact of a
leader fault in MultiPaxos by implementing a distributed leader, Cheap
Paxos~\cite{LamportMassa04CheapPaxos} optimizes Paxos to decrease the load
on some replicas in case failures are sufficiently separated in time,
Mencius~\cite{MaoJunqueiraMarzullo08MenciusBuildingEfficientReplicatedStateMachine} 
optimizes MultiPaxos for wide-area networks by rotating the role of leader
among all replicas, etc. Close scrutiny reveals that most SMR algorithms can
be seen as variations and optimizations around a few core ideas such as the
round structure introduced by Paxos~\cite{Lamport98ParttimeParliament}, quorum
systems~\cite{GuerraouiVukolic10RefinedQuorumSystems,Lamport06FastPaxos},
and agreement up to commutativity using command-structure
sets~\cite{Lamport05GeneralizeConsensus}.

Given their similarities, one may ask why is it not possible to reuse formal
analysis of SMR algorithms. The problem is that SMR algorithms are not modular
and that, because of their intricate structure, modifying a monolithic SMR
algorithm may produce many unforeseen corner cases that invalidate previous
analyses.

To simplify the task of devising new SMR algorithms we propose a framework
for building reusable SMR modules. For example, in \cref{}, we augment
MultiPaxos with our composition interface and prove that it satisfies Composable
Generalized Consensus (CGC), our correctness condition for modules. We then
build an SMR module whose structure is similar to the one of the fast rounds of
Fast Paxos and prove that it satisfies CGC\@. Our Fast Paxos module has lower
latency than our Paxos module when there is low contention among clients, but
its performance drops below that of Paxos when contention increases. Finally,
we assemble unmodified our MultiPaxos SMR module and our Fast Paxos SMR module
in an SMR algorithm that dynamically switch from one to the other depending on
the contention level observed in the system, therefore getting the best of both.
Crucially, we obtain the correctness of our composite SMR algorithm for free:
our generic composition theorem and the correctness of our two SMR modules taken
individually imply that the combination of the MultiPaxos SMR module and the
Fast Paxos SMR module is a correct SMR algorithm. Later, in \cref{}, we present
three other SMR modules that can be combined the MultiPaxos SMR module and the
Fast Paxos SMR module in a composite SMR algorithm that can switch between 5
different kinds of SMR modules, according to the system's operating conditions.

An SMR module is an SMR algorithm augmented with a \emph{composition interface}
and satisfying our correctness condition, called Composable Generalized
Consensus (CGC). The composition interface consists of an \emph{abort interface}
for aborting an SMR module's execution and passing the baton to another
SMR module and with a matching \emph{init interface} for taking over the
execution of an aborting SMR module. Two SMR modules are composed by passing
the outputs produced by the abort interface of the first SMR module, called
abort c-structs, as inputs to the init interface of the second SMR module,
called init c-structs in the context of the second SMR module. A SMR module
must satisfy a correctness condition that we call Composable Generalized
Consensus (abbreviated CGC). Any correct SMR module is a safe Generalized
Consensus~\cite{Lamport05GeneralizeConsensus} algorithm (i.e.\ a kind of SMR
algorithm), but the progress guarantees may vary between SMR modules: CGC only
requires a SMR module to make progress executing commands or to abort its
execution. CGC also relates the init c-structs, learned values (learned values
represent the local state-machine state of individual replicas), and abort
values in order for our composition theorem to hold.
%Note that Generalized Consensus is a specification of SMR algorithms that takes into account the commutativity relation among commands and can offer important performance advantages in practice.

Our composition theorem enables incremental development of SMR algorithms
because it guarantees that any number of correct modules may be composed in
a sequence to form a correct SMR algorithm. Therefore, our framework allows
building \emph{dynamic SMR algorithms}, which dynamically change module when
the currently running module aborts, choosing the module best suited to the
current operating conditions. Thanks to the composition theorem, new modules can
be added to a dynamic SMR algorithm without modifying the existing modules and
with guaranteed correctness. Moreover, the relaxed progress requirements of CGC
enable a separation of concerns that simplifies optimizing SMR algorithms for
particular scenarios because there is no need to consider executing commands
under a situation outside the scenario: the module can just abort in this case.

In practice, we provide a TLA+ specification of CGC that the user of our framework can use to debug and prove new modules correct using the TLA$^+$ toolbox~\cite{TLATools}.
Moreover, we also provide two TLA+ specifications simplifying the development of modules based on Refined Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems}. These two specifications both implement CGC and provide a skeleton for a user to build her own module. The two specifications demonstrate how to update the local state of a replica after gathering the states or the init c-structs of a write quorum of replicas, and how to
produce an abort c-struct after gathering the state of a read quorum of replicas. Each of the specifications achieves a particular consistency/latency trade-off~\cite{Abadi12ConsistencyTradeoffsModernDistributedDatabaseSystem}.
The two specifications use different combinations of sizes of read and write quorums, which affects the invariants that the replicas must maintain and, in consequence, the latency with which a command can be processed but the module.
Our two specifications abstract over how to disseminate the commands to the replicas (e.g.\ whether to use a leader as in Paxos, multiple leaders as in MultiCoordinatedPaxos, or whether to broadcast commands without using a leader as in Fast Paxos).
The user can therefore specify a concrete command dissemination procedure and then model-check that the obtained algorithm refines one of the two specifications.  
We demonstrate the practical use of our framework by presenting modules based on Paxos~\cite{Lamport98ParttimeParliament}, Fast Paxos\cite{Lamport06FastPaxos}, Chain Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability}, a case of Vertical Paxos~\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}, and EPaxos~\cite{MoraruAndersenKaminsky13ThereIsMoreConsensusEgalitarianParliaments}
(TODO, but would be nice).

Compared to related work, described in \cref{sec:related}, our key contributions are the following.
\begin{itemize}
    \item We present the first framework to allowing to build Generalized Consensus algorithms our of independent modules.
    \item We provide tool support for designing CGC modules in the form of two TLA+~\cite{Lamport02SpecifyingSystems} specifications simplifying the design and model-checking of SMR algorithms based on Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems} and demonstrate their use by designing CGC modules inspired by Paxos, Fast Paxos, and Primary-Backup replication.
\end{itemize}
These two contributions set this paper apart from related work, notably the Abstract framework~\cite{GuerraouiETAL10Next700BftProtocols} and the Speculative Linearizability framework~\cite{GuerraouiKuncakLosa12SpeculativeLinearizability}, both described in \cref{sec:related}.
Our TLA+ specifications can be found at \url{http://losa.fr/cgc}.
Moreover, we have also formalized the CGC correctness condition and proved the composition theorem in Isabelle/HOL\@, and the Isabelle theories are available at the same URL\@.

>>>>

Surprisingly, at a first glance, many SMR algorithms seem to share a common
structure which would indicate that previous formal analysis could be reused.
This similarity in structure exists because, in many cases, researchers obtained
a new SMR algorithm by optimizing an existing SMR algorithm for a particular use
case. For example, Fast Paxos~\cite{Lamport06FastPaxos} adds fast rounds
to MultiPaxos~\cite{lamport2001paxos} to reduce the latency under low
contention, Disk Paxos~\cite{GafniLamport03DiskPaxos} optimizes Paxos
for systems where disk nodes and processor nodes are distinct, Chain
Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability} 
organizes the replicas of Paxos in a ring to distribute
commands at a higher throughput but at the expense of latency,
FGGC~\cite{SutraShapiro11FastGenuineGeneralizedConsensus} optimizes
Generalized Paxos\cite{Lamport05GeneralizeConsensus} to reduce latency
when the system contains a reliable and fast leader node, Ring
Paxos~\cite{MarandiETAL10RingPaxosHighthroughputAtomicBroadcastProtocol}
optimizes the latency and throughput of Paxos when network multicast is
available, Multicoordinated Paxos reduces the performance impact of a
leader fault in MultiPaxos by implementing a distributed leader, Cheap
Paxos~\cite{LamportMassa04CheapPaxos} optimizes Paxos to decrease the load
on some replicas in case failures are sufficiently separated in time,
Mencius~\cite{MaoJunqueiraMarzullo08MenciusBuildingEfficientReplicatedStateMachine} 
optimizes MultiPaxos for wide-area networks by rotating the role of leader
among all replicas, etc. Close scrutiny reveals that most SMR algorithms can
be seen as variations and optimizations around a few core ideas such as the
round structure introduced by Paxos~\cite{Lamport98ParttimeParliament}, quorum
systems~\cite{GuerraouiVukolic10RefinedQuorumSystems,Lamport06FastPaxos},
and agreement up to commutativity based on command-structures~\cite{Lamport05GeneralizeConsensus}.

Given the similarities shared by many SMR algorithms, one may ask why is it
not possible to reuse formal analyses to simplify the verification of new
algorithms. The problem is that SMR algorithms are not modular and that, because
of their intricate structure, modifying a monolithic SMR algorithm may produce
many unforeseen corner cases that invalidate previous analyses.

To simplify the task of devising new SMR algorithms we propose a framework for
building self-contained and reusable SMR modules whose analysis is independent
of the other modules they may be composed with. For example, in \cref{}, we
augment MultiPaxos with our composition interface, and we prove that it satisfies
our correctness condition for modules, called Composable Generalized Consensus (CGC).
We then build an SMR module whose structure is similar to the one of the fast
rounds of Fast Paxos, and we prove that it satisfies CGC\@. Finally, we assemble
unmodified our MultiPaxos SMR module and our Fast Paxos SMR module in an SMR
algorithm that dynamically switches from one to the other depending on the
contention level observed in the system. This composite algorithm combines the
advantages of both its SMR modules: in cases of low contention, our Fast Paxos
module executes commands in two message delays, while MultiPaxos would need
three, whereas the MultiPaxos module maintains its throughput in high contention
cases, in which the performance of the Fast Paxos module would have been
significantly reduced.

Crucially, we obtain the correctness of our composite
SMR algorithm for free: our generic composition theorem and the correctness
of our two SMR modules taken individually imply that the combination of the
MultiPaxos SMR module and the Fast Paxos SMR module is a correct SMR algorithm.
Later, in \cref{}, we present three other SMR modules that can be combined the
MultiPaxos SMR module and the Fast Paxos SMR module in a composite SMR algorithm
that can switch between 4 different kinds of SMR modules, according to the
system's operating conditions.

More generally, an SMR module is an SMR algorithm augmented with a \emph{composition interface}
and satisfying a correctness condition that we call \emph{Composable Generalized
Consensus}, abbreviated CGC\@. The composition interface consists of an \emph{abort interface}
for aborting an SMR module's execution and passing the baton to another
SMR module and with a matching \emph{init interface} for taking over the
execution of an aborting SMR module. Two SMR modules are composed by passing
the outputs produced by the abort interface of the first SMR module, called
\emph{abort c-structs}, as inputs to the init interface of the second SMR module,
called \emph{init c-structs} in the context of the second SMR module. Any correct SMR module is a safe Generalized
Consensus~\cite{Lamport05GeneralizeConsensus} algorithm (i.e.\ a kind of SMR
algorithm), but the progress guarantees may vary between SMR modules: CGC only
requires an SMR module to make progress executing commands or to abort its
execution. CGC also relates the init c-structs, learned values (learned c-structs 
represent the local state-machine state of individual replicas), and abort
c-structs in order to guarantee that an SMR module consistently takes over the execution of its predecessor, as expressed in our composition theorem.
Finally, our framework allows building \emph{dynamic SMR
algorithms}, which dynamically change module when the currently running module
aborts, choosing the module best suited to the current operating conditions.

Our composition theorem enables incremental development of SMR algorithms
because it guarantees that any number of correct modules may be composed in a
sequence to form a correct SMR algorithm. Thanks to the composition theorem,
new modules can be added to a dynamic SMR algorithm without modifying the
existing modules and with guaranteed correctness. Moreover, the relaxed progress
requirements of CGC enable a separation of concerns that simplifies optimizing
SMR algorithms for particular scenarios because there is no need to consider
executing commands under a situation outside the scenario: the module can just
abort in this case.

In practice, we provide a TLA+ specification of CGC that the user of our framework can use to debug and prove new modules correct using the TLA$^+$ toolbox~\cite{TLATools}.
Moreover, we also provide two TLA+ specifications simplifying the development of modules based on Refined Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems}. These two specifications both implement CGC and provide a skeleton for a user to build her own module. The two specifications demonstrate how to update the local state of a replica after gathering the states or the init c-structs of a write quorum of replicas, and how to
produce an abort c-struct after gathering the state of a read quorum of replicas. Each of the specifications achieves a particular consistency/latency trade-off~\cite{Abadi12ConsistencyTradeoffsModernDistributedDatabaseSystem}.
The two specifications use different combinations of sizes of read and write quorums, which affects the invariants that the replicas must maintain and, in consequence, the latency with which a command can be processed but the module.
Our two specifications abstract over how to disseminate the commands to the replicas (e.g.\ whether to use a leader as in Paxos, multiple leaders as in MultiCoordinatedPaxos, or whether to broadcast commands without using a leader as in Fast Paxos).
The user can therefore specify a concrete command dissemination procedure and then model-check that the obtained algorithm refines one of the two specifications.  
We demonstrate the practical use of our framework by presenting modules based on Paxos~\cite{Lamport98ParttimeParliament}, Fast Paxos\cite{Lamport06FastPaxos}, Chain Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability}, a case of Vertical Paxos~\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}.

Compared to related work, described in \cref{sec:related}, our key contributions are the following.
\begin{itemize}
    \item We present the first framework allowing to build Generalized Consensus algorithms our of independent and reusable modules supporting modular formal analysis.
    \item We provide tool support for designing CGC modules in the form of two TLA+~\cite{Lamport02SpecifyingSystems} specifications simplifying the design and model-checking of SMR algorithms based on Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems} and demonstrate their use by designing four CGC modules inspired by MultiPaxos, Fast Paxos, Chain Replication, and Primary-Backup replication.
\end{itemize}
These two contributions set this paper apart from related work, notably the Abstract framework~\cite{GuerraouiETAL10Next700BftProtocols} and the Speculative Linearizability framework~\cite{GuerraouiKuncakLosa12SpeculativeLinearizability}, both described in \cref{sec:related}.
Our TLA+ specifications can be found at \url{http://losa.fr/cgc}.
Moreover, we have also formalized the CGC correctness condition and proved the composition theorem in Isabelle/HOL\@, and the Isabelle theories are available at the same URL\@.
